---
title: "Data Visualization"
author: "Yufei Liu"
output: 
    html_document:
        toc: TRUE
        toc_float: TRUE
---

```{r setup, message=FALSE, include=FALSE}
library(data.table)
library(plotly)
library(ggplot2)
library(ggstats)
```

```{r read-data, include=FALSE}
# Import the dataset
data_vis <- read.csv("./data/vis_data.csv")
```

Below are the data visualization techniques used in exploratory data analysis after preprocessing the dataset.

## Proportional Barplots of the Proportion of Having Diabetes in Different Categories of the Categorical Variables {.tabset}

```{r plot4, class.source="code-r-small", include=FALSE}
# Generate plot 4
p4_bar <- ggplot(data_vis, aes(x = factor(race), fill = diabetes, by = factor(race))) + 
  geom_bar(position = "fill", alpha = 0.75) + 
  labs(
    title = "Plot 4: Proportion of Diabetes in Different Races", 
    x = "Race", 
    y = "Proportion",
    fill = "Diabetes"
    ) +
  geom_text(stat = "prop", position = position_fill(.5)) +
  theme(plot.title = element_text(size=10)) +
  coord_flip()

```

```{r plot5, class.source="code-r-small", include=FALSE}
# Generate plot 5
p5_bar <- ggplot(data_vis, aes(x = factor(sex), fill = diabetes, by = factor(sex))) + 
  geom_bar(position = "fill", alpha = 0.75) + 
  labs(
    title = "Plot 5: Proportion of Diabetes in Different Sexes", 
    x = "Sex", 
    y = "Proportion",
    fill = "Diabetes"
    ) +
  geom_text(stat = "prop", position = position_fill(.5)) +
  theme(plot.title = element_text(size=10)) +
  coord_flip()

```

```{r plot6, class.source="code-r-small", include=FALSE}
# Generate plot 6
education_order <- c("Did not graduate High School", "Graduated High School", "Attended College or Technical School", "Graduated from College or Technical School")
  
p6_bar <- ggplot(data_vis, aes(x = factor(education, level = education_order), fill = diabetes, by = factor(education, level = education_order))) + 
  geom_bar(position = "fill", alpha = 0.75) + 
  labs(
    title = "Plot 6: Proportion of Diabetes in People with Different Education Level", 
    x = "Education Level", 
    y = "Proportion",
    fill = "Diabetes"
    ) +
  geom_text(stat = "prop", position = position_fill(.5)) +
  theme(plot.title = element_text(size=6)) +
  coord_flip()

```

```{r plot7, class.source="code-r-small", include=FALSE}
# Generate plot 7
p7_bar <- ggplot(data_vis, aes(x = factor(sleep), fill = diabetes, by = factor(sleep))) + 
  geom_bar(position = "fill", alpha = 0.75) + 
  labs(
    title = "Plot 7: Proportion of Diabetes in People with Different Sleep Hours", 
    x = "Healthy Sleep", 
    y = "Proportion",
    fill = "Diabetes"
    ) +
  geom_text(stat = "prop", position = position_fill(.5)) +
  theme(plot.title = element_text(size=8)) +
  coord_flip()

```

```{r plot8, class.source="code-r-small", include=FALSE}
# Generate plot 8
p8_bar <- ggplot(data_vis, aes(x = factor(smoke), fill = diabetes, by = factor(smoke))) + 
  geom_bar(position = "fill", alpha = 0.75) + 
  labs(
    title = "Plot 8: Proportion of Diabetes in People with Different Smoke Habits", 
    x = "Smoke", 
    y = "Proportion",
    fill = "Diabetes"
    ) +
  geom_text(stat = "prop", position = position_fill(.5)) +
  theme(plot.title = element_text(size=8)) +
  coord_flip()

```

```{r plot9, class.source="code-r-small", include=FALSE}
# Generate plot 9
p9_bar <- ggplot(data_vis, aes(x = factor(exercise), fill = diabetes, by = factor(exercise))) + 
  geom_bar(position = "fill", alpha = 0.75) + 
  labs(
    title = "Plot 9: Proportion of Diabetes in People with Different Exercise Habits", 
    x = "Exercise", 
    y = "Proportion",
    fill = "Diabetes"
    ) +
  geom_text(stat = "prop", position = position_fill(.5)) +
  theme(plot.title = element_text(size=10)) +
  coord_flip()

```

```{r plot10, class.source="code-r-small", include=FALSE}
# Generate plot 10
p10_bar <- ggplot(data_vis, aes(x = factor(stroke), fill = diabetes, by = factor(stroke))) + 
  geom_bar(position = "fill", alpha = 0.75) + 
  labs(
    title = "Plot 10: Proportion of Diabetes in People with Stroke or Not", 
    x = "Stroke", 
    y = "Proportion",
    fill = "Diabetes"
    ) +
  geom_text(stat = "prop", position = position_fill(.5)) +
  theme(plot.title = element_text(size=10)) +
  coord_flip()

```

```{r plot11, class.source="code-r-small", include=FALSE}
# Generate plot 11
p11_bar <- ggplot(data_vis, aes(x = factor(heart_disease), fill = diabetes, by = factor(heart_disease))) + 
  geom_bar(position = "fill", alpha = 0.75) + 
  labs(
    title = "Plot 11: Proportion of Diabetes in People with Heart Diseases or Not", 
    x = "Heart Disease", 
    y = "Proportion",
    fill = "Diabetes"
    ) +
  geom_text(stat = "prop", position = position_fill(.5)) +
  theme(plot.title = element_text(size=10)) +
  coord_flip()

```

```{r plot12, class.source="code-r-small", include=FALSE}
# Generate plot 12
p12_bar <- ggplot(data_vis, aes(x = factor(alcohol), fill = diabetes, by = factor(alcohol))) + 
  geom_bar(position = "fill", alpha = 0.75) + 
  labs(
    title = "Plot 12: Proportion of Diabetes in People with Different Alcohol Habits", 
    x = "Alcohol", 
    y = "Proportion",
    fill = "Diabetes"
    ) +
  geom_text(stat = "prop", position = position_fill(.5)) +
  theme(plot.title = element_text(size=10)) +
  coord_flip()

```

```{r plot13, class.source="code-r-small", include=FALSE}
# Generate plot 13
health_order <- c("Poor", "Fair", "Good", "Very good", "Excellent")
p13_bar <- ggplot(data_vis, aes(x = factor(health, level = health_order), fill = diabetes, by = factor(health, level = health_order))) + 
  geom_bar(position = "fill", alpha = 0.75) + 
  labs(
    title = "Plot 13: Proportion of Diabetes in People with Different Health Levels", 
    x = "Health Level", 
    y = "Proportion",
    fill = "Diabetes"
    ) +
  geom_text(stat = "prop", position = position_fill(.5)) +
  theme(plot.title = element_text(size=10)) +
  coord_flip()

```

### Plot 4: Race

```{r echo=FALSE, warning=FALSE}
# Interactive Plot 4
ggplotly(p4_bar)
```

Plot 4 shows the proportion of people with diabetes among various races.  
There are more people without diabetes than with diabetes in most races except for Black and American Indian/Alaskan Native. The proportion of having diabetes is relatively high in American Indian/Alaskan Native (75.5%) and Black (69.9%), and relatively low in Asian (17.5%) and Hispanic (25.0%).

### Plot 5: Sex

```{r echo=FALSE, warning=FALSE}
# Interactive Plot 5
ggplotly(p5_bar)
```

Plot 5 shows the proportion of people with diabetes in male and female.  
The proportions are roughly the same among two sex, with the proportion of diabetes in male (50.7%) slightly higher than in female (49.3%).

### Plot 6: Education

```{r echo=FALSE, warning=FALSE}
# Interactive Plot 6
ggplotly(p6_bar)
```

Plot 6 shows the proportion of diabetes among people with different education level.  
As the education level increases, the proportion decreases, which indicates a negative relationship between education level and diabetes.

### Plot 7: Sleep

```{r echo=FALSE, warning=FALSE}
# Interactive Plot 7
ggplotly(p7_bar)
```

Plot 7 shows the proportion of diabetes among people who sleep 7 to 10 hours (inclusive) on average in a day and people who sleep less than 7 hours or more than 10 hours on average in a day.  
The proportion of diabetes is slightly higher in the latter group (53.3%) than the prior group (48.2%). This indicates a negative relationship between appropriate sleep and diabetes.

### Plot 8: Smoke

```{r echo=FALSE, warning=FALSE}
# Interactive Plot 8
ggplotly(p8_bar)
```

Plot 8 shows the proportion of people with diabetes among people who smoked more than 100 cigarettes in their entire life and people who didn't.  
The proportion of people with diabetes among heavy smokers (55.2%) are higher than the proportion of people who didn't smoke much (45.7%). This indicates a positive relationship between smoking a lot and having diabetes.

### Plot 9: Exercise

```{r echo=FALSE, warning=FALSE}
# Interactive Plot 9
ggplotly(p9_bar)
```

Plot 9 shows the proportion of people with diabetes among people who exercised in the past 30 days and people who didn't.  
The proportion of people with diabetes among exercised people (43.3%) are much lower than the proportion of people who didn't exercise recently (66.1%). This indicates a negative relationship between exercise and having diabetes.

### Plot 10: Stroke

```{r echo=FALSE, warning=FALSE}
# Interactive Plot 10
ggplotly(p10_bar)
```

Plot 10 shows the proportion of people with diabetes among people who ever had stroke before and people who didn't.  
The proportion of people with diabetes among people that had stroke (75.7%) are much higher than the proportion of people who didn't have stroke (48.1%). This indicates a positive relationship between stroke and diabetes.

### Plot 11: Heart Disease

```{r echo=FALSE, warning=FALSE}
# Interactive Plot 11
ggplotly(p11_bar)
```

Plot 11 shows the proportion of people with diabetes among people who ever had coronary heart disease (CHD) or myocardial infarction (MI) before and people who didn't.  
The proportion of people with diabetes among people that had heart disease (75.9%) are much higher than the proportion of people who didn't have stroke (45.4%). This indicates a positive relationship between heart disease and diabetes.

### Plot 12: Alcohol

```{r echo=FALSE, warning=FALSE}
# Interactive Plot 12
ggplotly(p12_bar)
```

Plot 12 shows the proportion of people with diabetes among adult men having more than 14 drinks per week and adult women having more than 7 drinks per week and people who didn't.  
The proportion of people with diabetes among heavy drinkers (26.1%) are much lower than the proportion of people who aren't heavy drinkers (51.4%). This indicates a negative relationship between alcohol and diabetes, which is surprising.

### Plot 13: Health

```{r echo=FALSE, warning=FALSE}
# Interactive Plot 13
ggplotly(p13_bar)
```

Plot 13 shows the proportion of diabetes among people with different self-reported health level.  
As the health level increases, the proportion decreases, which indicates a negative relationship between health level and diabetes.
